---
title: "CITE clustering"
date: "`r Sys.Date()`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r knitr, include = FALSE}
source(here::here("R/setup_knitr.R"))
```

```{r setup, cache = FALSE}
#### LIBRARIES ####
# Package conflicts
library("conflicted")

# Single-cell
library("SingleCellExperiment")
library("scran")
library("scater")

# Plotting
library("clustree")
library("ggforce")

# Bioconductor
library("BiocSingular")

# File paths
library("fs")
library("here")

# Presentation
library("knitr")
library("jsonlite")

# Tidyverse
library("tidyverse")

### CONFLICT PREFERENCES ####
conflict_prefer("path", "fs")
conflict_prefer("mutate", "dplyr")

### SOURCE FUNCTIONS ####
source(here("R/output.R"))

### OUTPUT DIRECTORY ####
OUT_DIR <- here("output", DOCNAME)
dir_create(OUT_DIR)

#### SET GGPLOT THEME ####
theme_set(theme_minimal())

#### SET PATHS ####
source(here("R/set_paths.R"))
```

Introduction
============

In document we are going to perform clustering using just the CITE data. This 
analysis is based on the ["simpleSingleCell"][simple] workflow.

```{r load, cache.extra = c(tools::md5sum(PATHS$cite_qc))}
if (all(file_exists(c(PATHS$cite_qc)))) {
    cite <- read_rds(PATHS$cite_qc)
} else {
    stop("Filtered dataset is missing. ",
         "Please run '02-quality-control.Rmd' first.",
         call. = FALSE)
}

col_data <- as.data.frame(colData(cite))
```

Normalisation
=============

Because the CITE dataset is so much smaller we will just use library size
factors for normalisation instead of a more complex approach.

```{r norm}
sizeFactors(cite) <- librarySizeFactors(cite)
cite <- normalize(cite)
```

Mean-variance
=============

Fit
----

Fit the mean-variance relationship

```{r fit-mean-var}
tech_trend <- makeTechTrend(x = cite)
fit <- trendVar(cite, use.spikes = FALSE, loess.args = list(span = 0.5))

plot_data <- tibble(
    Mean = fit$means,
    Var = fit$vars,
    Trend = fit$trend(fit$means),
    TechTrend = tech_trend(fit$means)
)

ggplot(plot_data, aes(x = Mean)) +
    geom_point(aes(y = Var)) +
    geom_line(aes(y = Trend), colour = "blue") +
    geom_line(aes(y = TechTrend), colour = "red") +
    annotate("text", x = 7, y = 2, label = "Fitted trend", colour = "blue") +
    annotate("text", x = 7, y = 1.8, label = "Poisson noise", colour = "red") +
    labs(
        title = "Mean-variance relationship",
        x = "Mean log-expression",
        y = "Variance of log-expression"
    )
```

Dimensionality reduction
========================

PCA
---

PCA is performed on the dataset and components selected based on the proportion
of variance explained.

```{r denoise}
set.seed(1)
cite <- runPCA(cite, ncomponents = 30, BSPARAM = IrlbaParam())
n_pcs <- 10

plot_data <- tibble(
    PC = seq_along(attr(reducedDim(cite), "percentVar")),
    PercentVar = attr(reducedDim(cite), "percentVar")
) %>%
    mutate(Selected = PC <= n_pcs)

ggplot(plot_data, aes(x = PC, y = PercentVar, colour = Selected)) +
    geom_point() +
    scale_colour_manual(values = c("grey40", "red")) +
    labs(
        title = "PC variance",
        x = "Principal component",
        y = "Proportion of variance explained"
    )

plotPCA(cite, ncomponents = 3, colour_by = "total_features_by_counts")
```

Here we select the first `r n_pcs` components.

t-SNE
-----

```{r t-SNE}
set.seed(1)
cite <- runTSNE(cite, use_dimred = "PCA", perplexity = 40)
plotTSNE(cite, colour_by = "log10_total_counts")
```

Clustering
==========

Cluster cells using the implementation in `Seurat` because it has a resolution
parameter.

```{r clustering, results = "hide"}
snn_mat <- Seurat::FindNeighbors(reducedDim(cite, "PCA"))$snn

resolutions <- seq(0, 1, 0.1)
for (res in resolutions) {
    clusters <- Seurat:::RunModularityClustering(snn_mat, resolution = res)
    col_data[[paste0("ClusterRes", res)]] <- factor(clusters)
}

colData(cite) <- DataFrame(col_data)
```

Dimensionality reduction {.tabset}
------------------------

Dimensionality reduction plots showing clusters at different resolutions.

### PCA {.tabset}

```{r pca-res, results = "hide"}
src_list <- lapply(resolutions, function(res) {
    src <- c(
        "#### Res {{res}} {.unnumbered}",
        "```{r res-pca-{{res}}}",
        "plotPCA(cite, colour_by = 'ClusterRes{{res}}') + theme_minimal()",
        "```",
        ""
    )
    knit_expand(text = src)
})
out <- knit_child(text = unlist(src_list), options = list(cache = FALSE))
```

`r out`

### t-SNE {.tabset}

```{r tSNE-res, results = "hide"}
src_list <- lapply(resolutions, function(res) {
    src <- c(
        "#### Res {{res}} {.unnumbered}",
        "```{r res-tSNE-{{res}}}",
        "plotTSNE(cite, colour_by = 'ClusterRes{{res}}') + theme_minimal()",  
        "```",
        ""
    )
    knit_expand(text = src)
})
out <- knit_child(text = unlist(src_list), options = list(cache = FALSE))
```

`r out`

Clustering trees {.tabset}
----------------

Clustering trees show the relationship between clusterings at adjacent
resolutions. Each cluster is represented as a node in a graph and the edges show
the overlap between clusters.

### Standard

Coloured by clustering resolution.

```{r clustree}
clustree(cite, prefix = "ClusterRes")
```

### Stability

Coloured by the SC3 stability metric.

```{r clustree-stability}
clustree(cite, prefix = "ClusterRes", node_colour = "sc3_stability")
```

Selection
---------

```{r select-res}
res <- 0.8
col_data$Cluster <- col_data[[paste0("ClusterRes", res)]]
colData(cite) <- DataFrame(col_data)
n_clusts <- length(unique(col_data$Cluster))
```

Based on these plots we will use a resolution of `r res` which gives us
`r n_clusts` clusters.

Validation {.tabset}
==========

To validate the clusters we will repeat some of our quality control plots
separated by cluster. At this stage we just want to check that none of the
clusters are obviously the result of technical factors.

Cluster {.tabset}
-------

Clusters assigned by `Seurat`.

### Count

```{r val-cluster-count}
ggplot(col_data, aes(x = Cluster, fill = Cluster)) +
    geom_bar()
```

### PCA

```{r val-cluster-pca}
plotPCA(cite, colour_by = "Cluster", point_alpha = 1) +
    scale_fill_discrete() +
    theme_minimal()
```

### t-SNE

```{r val-cluster-tSNE}
plotTSNE(cite, colour_by = "Cluster", point_alpha = 1) +
    scale_fill_discrete() +
    theme_minimal()
```

Total counts {.tabset}
------------

Total counts per cell.

### Distribution

```{r val-counts-dist}
ggplot(col_data, aes(x = Cluster, y = log10_total_counts, colour = Cluster)) +
    geom_violin() +
    geom_sina(size = 0.5) +
    theme(legend.position = "none")
```

### PCA

```{r val-counts-pca}
plotPCA(cite, colour_by = "log10_total_counts", point_alpha = 1) +
    scale_fill_viridis_c() +
    theme_minimal()
```

### t-SNE

```{r val-counts-tSNE}
plotTSNE(cite, colour_by = "log10_total_counts", point_alpha = 1) +
    scale_fill_viridis_c()+
    theme_minimal()
```

Total features {.tabset}
--------------

Total number of expressed features per cell.

### Distribution

```{r val-features-dist}
ggplot(col_data,
       aes(x = Cluster, y = total_features_by_counts, colour = Cluster)) +
    geom_violin() +
    geom_sina(size = 0.5) +
    theme(legend.position = "none")
```

### PCA

```{r val-features-pca}
plotPCA(cite, colour_by = "total_features_by_counts", point_alpha = 1) +
    scale_fill_viridis_c() +
    theme_minimal()
```

### t-SNE

```{r val-features-tSNE}
plotTSNE(cite, colour_by = "total_features_by_counts", point_alpha = 1) +
    scale_fill_viridis_c()+
    theme_minimal()
```

Summary
=======

Parameters
----------

This table describes parameters used and set in this document.

```{r parameters}
params <- list(
    list(
        Parameter = "n_pcs",
        Value = n_pcs,
        Description = "Selected number of principal components for clustering"
    ),
    list(
        Parameter = "resolutions",
        Value = resolutions,
        Description = "Range of possible clustering resolutions"
    ),
    list(
        Parameter = "res",
        Value = res,
        Description = "Selected resolution parameter for clustering"
    ),
    list(
        Parameter = "n_clusts",
        Value = n_clusts,
        Description = "Number of clusters produced by selected resolution"
    )
)

params <- toJSON(params, pretty = TRUE)
kable(fromJSON(params))
```

Output files
------------

This table describes the output files produced by this document. Right click
and _Save Link As..._ to download the results.

```{r output}
write_rds(cite, PATHS$cite_clust, compress = "bz", compression = 9)
write_lines(params, path(OUT_DIR, "parameters.json"))

kable(data.frame(
    File = c(
        download_link("parameters.json", OUT_DIR)
    ),
    Description = c(
        "Parameters set and used in this analysis"
    )
))
```

Session information
-------------------

[simple]: https://bioconductor.org/packages/release/workflows/html/simpleSingleCell.html
